Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-04-01 vertical-thermal-review.Rmd
merge two paragraphs in warming section
#113

#Word count
w_intro=414
w_biophys=1580
w_tleaf=758
w_traits= 2477
w_met=1000
w_ecol=1352
w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
w_implications=2399
w_conclusions=248
w_total=w_intro+w_vertgradients+w_implications+w_conclusions

@@

<!--the following two paragraphs could be shortened/ condensed into one-->
It remains poorly understood how responses to warming will vary across vertical gradients under mesic conditions. 
Leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to local climate but do not vary consistently with height (see section 4.2, Table 2). 
Beyond the optimal temperature, photosynthesis declines whereas respiration continues to increase, shifting the carbon balance from source to sink, independent of water and irradiance [@duffyHowCloseAre2021]. 
While leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will reduce carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019]. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].

Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
<!--Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (see section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].

>>>>

#Word count
w_intro=414
w_biophys=1580
w_tleaf=758
w_traits= 2477
w_met=1000
w_ecol=1352
w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
w_implications=2333
w_conclusions=248
w_total=w_intro+w_vertgradients+w_implications+w_conclusions

@@

It remains poorly understood how responses to warming will vary across vertical gradients under mesic conditions. 
Leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], yet failure to fully acclimate will reduce carbon sequestration at leaf, tree, and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
As growing season $T_{air}$ increasingly equals or exceeds photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], exposed crowns are likely most vulnerable. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and $T_{crit}$ [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress]. 
In contrast, understory leaves and trees are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), but in some ways are likely to prove more sensitive to higher $T_{air}$.
Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
<!--Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (see section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].
